---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(rsimd)
```

# rsimd

<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
[![R-CMD-check](https://github.com/coolbutuseless/rsimd/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/coolbutuseless/rsimd/actions/workflows/R-CMD-check.yaml)
<!-- badges: end -->


`{rsimd}` is a demonstration package showing how cross-platform [SIMD](https://en.wikipedia.org/wiki/Single_instruction,_multiple_data)
instructions can be used in the compiled C code in a package.

This package makes use of [SIMDe](https://github.com/simd-everywhere/simde) - 
a header-only C library which provides a SIMD translation layer such that
SIMD intrinsics from one CPU architecture may be run on a different CPU 
archtecture e.g. running AVX512 instructions on ARM Neon.

In this package, SIMD instructions are written in [x86 AVX512](https://en.wikipedia.org/wiki/AVX-512)
[intrinsics](https://en.wikipedia.org/wiki/Intrinsic_function) and should run natively on modern x86 CPUs.   

By using SIMDe library, these x86 AVX512
automatically uses [ARM Neon](https://en.wikipedia.org/wiki/ARM_architecture_family#Advanced_SIMD_(Neon)) SIMD 
instructions in order to run fast on Mac AArch64 ARM processors.


### What's in the box


* `cpuid()` output some very basic CPU ID information about which SIMD 
  instructions sets your CPU supports.
* Four implementations of a fused muliply-add
    * `multiply_add_r()` Pure R code which just runs `a * b + c`
    * `multiply_add_c()` Naive C code which does the same operation in C
    * `multiply_add_unrolled()` C code in which the inner for-loop is unrolled for performance
    * `multiply_add_simde()` C code including AVX512 SIMD instructions 



## Installation

You can install from [GitHub](https://github.com/coolbutuseless/rsimd) with:

``` r
# install.packages('remotes')
remotes::install_github('coolbutuseless/rsimd')
```


#### OS Notes

The SIMD code should compile and work on both x86 and mac ARM processors.

Please let me know if this is not the case for your machine.



## CPU ID

Print some basic notes about your CPU and the supported SIMD instructions.

```{r}
cpuid()
```


## Included case study: Fused multiply-add

The case-study explored in this package is to speed up `fused multiply add` in R by 
using SIMD instructions

A common operation is to multiply two vectors and add a constant e.g. `a * b + c`.

In SIMD, this is sometimes referred to as a 'fused multiply add'.

The following code benchmarks 4 implementations 

General findings on my mac ARM M2 CPU

* Plain R implementation is faster than C
* Unrolled C code is faster than R
* SIMD code is faster than all


```{r example}
library(rsimd)

N <- 10000
a <- runif(N)
b <- runif(N)
c <- runif(N)

bench::mark(
  multiply_add_r(a, b, c),
  multiply_add_c(a, b, c),
  multiply_add_unrolled(a, b, c),
  multiply_add_simde(a, b, c)
)[, 1:5] |>
  knitr::kable()
```




## R and C code implementations

#### R code from `R/simd.R`

<details>
<summary>`multiply_add_r()`</summary>

``` r
multiply_add_r <- function(a, b, c) {
  stopifnot(length(a) == length(b) && length(a) == length(c))
  a * b + c
}
```

</details>


#### The following C code sections are from this package `src/simd.c`

<details>
<summary>`multiply_add_c()`</summary>

``` c
SEXP multiply_add_c_(SEXP a_, SEXP b_, SEXP c_) {

  if (length(a_) != length(b_) || length(a_) != length(c_)) {
    error("All arguments must be the same length");
  }
  
  
  double *a = REAL(a_);
  double *b = REAL(b_);
  double *c = REAL(c_);
  
  SEXP res_ = PROTECT(allocVector(REALSXP, length(a_)));
  double *res = REAL(res_);
  
  for (int i = 0; i < length(a_); i++) {
    res[i] = a[i] * b[i] + c[i];
  }
  
  
  UNPROTECT(1);
  return res_;
}
```

</details>

<details>
<summary>`multiply_add_unrolled()`</summary>

```
SEXP multiply_add_unrolled_(SEXP a_, SEXP b_, SEXP c_) {
  
  if (length(a_) != length(b_) || length(a_) != length(c_)) {
    error("All arguments must be the same length");
  }
  
  
  double *a = REAL(a_);
  double *b = REAL(b_);
  double *c = REAL(c_);
  
  SEXP res_ = PROTECT(allocVector(REALSXP, length(a_)));
  double *res = REAL(res_);

#define UNROLL 8 
  int i = 0;
  for (; i < length(a_) - (UNROLL - 1); i += UNROLL) {
    *res++ = *a++ * *b++ + *c++;
    *res++ = *a++ * *b++ + *c++;
    *res++ = *a++ * *b++ + *c++;
    *res++ = *a++ * *b++ + *c++;
    
    *res++ = *a++ * *b++ + *c++;
    *res++ = *a++ * *b++ + *c++;
    *res++ = *a++ * *b++ + *c++;
    *res++ = *a++ * *b++ + *c++;
  }
  for (; i< length(a_); i++) {
    res[i] = a[i] * b[i] + c[i];
  }
  
  
  UNPROTECT(1);
  return res_;
}
```


</details>

<details>
<summary>`multiply_add_simde()`</summary>

```
#define SIMDE_ENABLE_NATIVE_ALIASES 1
#include "simde/x86/avx512.h"

SEXP multiply_add_simde_(SEXP a_, SEXP b_, SEXP c_) {
  
  if (length(a_) != length(b_) || length(a_) != length(c_)) {
    error("All arguments must be the same length");
  }
  
  
  double *ap = REAL(a_);
  double *bp = REAL(b_);
  double *cp = REAL(c_);
  
  SEXP res_ = PROTECT(allocVector(REALSXP, length(a_)));
  double *resp = REAL(res_);
  
#define UNROLL 8 
  int i = 0;
  for (; i < length(a_) - (UNROLL - 1); i += UNROLL) {
    __m512d a = _mm512_load_pd(ap + i);
    __m512d b = _mm512_load_pd(bp + i);
    __m512d c = _mm512_load_pd(cp + i);
    __m512d res = _mm512_fmadd_pd(a, b, c);
    _mm512_store_pd(resp + i, res);
  }
  for (; i< length(a_); i++) {
    resp[i] = ap[i] * bp[i] + cp[i];
  }
  
  
  UNPROTECT(1);
  return res_;
}

```


</details>




## References

* [Intel's guide to SIMD intrinsics](https://www.intel.com/content/www/us/en/docs/intrinsics-guide/index.html)
* [ARM Neon Intrinsics Reference](file:///Users/mike/Desktop/video/advsimd-2021Q2.pdf)
* [SIMDe library](https://github.com/simd-everywhere/simde)
